Missing The Point: Non-Convergence in Iterative Imputation Algorithms

### 2021-09-17 Manuscript.Rmd
Add reference against visual inspection of several parameters
Save manuscript before editing

Iterative imputation has become the de facto standard to accommodate for missing data. The aim is usually to draw valid inferences, i.e. to get unbiased, confidence-valid estimates that incorporate the effects of the missingness. Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem. The missing values are imputed (i.e., filled in) using some sort of algorithm. And subsequently, the scientific model of interest is performed on the completed data. To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered. The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.

All inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial. There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to assess convergence would be preferred.*

@@

>>>>

Iterative imputation has become the de facto standard to accommodate missing data in social scientific research. The technique enables researchers to draw valid inferences despite having incomplete observations. Unbiased, confidence-valid estimates are obtained by separating the missing data problem from the scientific problem. First, the missing values are imputed (i.e., filled in) using some sort of algorithm. Only subsequently, the scientific model of interest is performed on the completed data. To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered. The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values.

The scientific inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial. There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to assess convergence would be preferred.*

@@

Add reference: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p. 285).
